---
title: "Introduction to ggplot2"
subtitle: Karthik Ram
date: "NHM"
output:
  xaringan::moon_reader:
   # css: "my-theme.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
class: inverse, center, middle

# Why ggplot2?

---
## ggplot2

-   Follows a grammar, just like any language.

-   It defines basic components that make up a sentence. In this case,
    the grammar defines components in a plot.
--

---

# Why not Base graphics?


-   Ugly, laborious, and verbose

-   There are better ways to describe statistical visualizations.


---

# Install ggplot2

```r
install.packages("ggplot2")
```

---

# Basics of ggplot2

### Some terminology

-   **ggplot** - The main function where you specify the dataset and
    variables to plot

-   **geoms** - geometric objects

    -   `geom_point()`, `geom_bar()`, `geom_density()`, `geom_line()`, `geom_area()`

-   **aes** - aesthetics

    -   shape, transparency (alpha), color, fill, linetype.

-   **scales** Define how your data will be plotted

    -   *continuous*, *discrete*, *log*
    
---

class: inverse, center, middle

"+" versus "%>%"

---

# Let's make a first ggplot2

```{r eval = FALSE}
head(iris)
View(head(iris))

iris %>% head %>% View
```




# Let's try an example

```{r, code, fig.height=4}
library(ggplot2)
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
geom_point()
```


---

# Basic structure of a ggplot2


```{r, code2, fig.height=4}
library(ggplot2)
ggplot(data = iris,
aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
```
---
# or build them interatively
```{r fig.height=4}  
myplot <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
myplot + geom_point()
```

---
# Add some color

```{r, fig.height=4}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
   geom_point(size = 3)
```

---

# Add some shape

```{r fig.height = 4}
ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(aes(shape = Species), size = 3)
```

---

# Exercise 1

Generate the plot below (15 minutes):

tip: `d2 <- sample_n(diamonds, 1000)`

```{r echo=FALSE, fig.height = 4}
d2 <- diamonds[sample(1:dim(diamonds)[1], 1000),]
ggplot(d2, aes(carat, price, color = color)) + geom_point() + theme_gray()
```

---

# Histograms are similarly easy

```{r, fig.height = 4}
h <- ggplot(faithful, aes(x = waiting))
h + geom_histogram(binwidth = 30, colour = "black")
```
---

Customize your histograms

```{r, fig.height=4}
h <- ggplot(faithful, aes(x = waiting))
h + geom_histogram(binwidth = 8,
fill = "steelblue",
colour = "black")
```

---

Let's read some climate data in:

```{r, fig.height = 4}
climate <-
  read.csv(
  "https://raw.githubusercontent.com/karthikram/ggplot-lecture/master/climate.csv",
  header = T
  )
  
  ggplot(climate, aes(Year, Anomaly10y)) +
  geom_line()
```

URL is `https://raw.githubusercontent.com/karthikram/ggplot-lecture/master/climate.csv`

---

We can also plot confidence regions


```{r, fig.height=4}
ggplot(climate, aes(Year, Anomaly10y)) +
  geom_ribbon(
  aes(ymin = Anomaly10y - Unc10y,
  ymax = Anomaly10y + Unc10y),
  fill = "blue",
  alpha = .1
  ) +
  geom_line(color = "steelblue")
```
---

Modify the previous plot and change it such that there are three lines instead of one with a confidence band. (Take 15 minutes)

```{r, fig.height=4, echo = FALSE}
cplot <- ggplot(climate, aes(Year, Anomaly10y))
cplot <- cplot + geom_line(size = 0.7, color = "black")
cplot <-
cplot + geom_line(
aes(Year, Anomaly10y + Unc10y),
linetype = "dashed",
size = 0.7,
color = "red"
)
cplot <-
cplot + geom_line(
aes(Year, Anomaly10y - Unc10y),
linetype = "dashed",
size = 0.7,
color = "red"
)
cplot + theme_gray()
```

Tips: You can keep adding geoms to an existing plot

You can add or subtract values from a column on the fly

---
# Bar plots

```{r fig.height=4}
ggplot(iris, aes(Species, Sepal.Length)) +
geom_bar(stat = "identity")
```
---

# A stacked bar

```{r fig.height = 4}
iris %>% gather(variable, value,-Species) %>%
  ggplot(aes(Species, value, fill = variable)) +
  geom_bar(stat = "identity")
```

---


# Let's try some other data : mammal survey data


```{r, eval = TRUE, cache = TRUE}
download.file("https://ndownloader.figshare.com/files/2292169","portal_data_joined.csv")
surveys <- read.csv('portal_data_joined.csv')
```

       
---

# Minor data cleanup

```{r echo = FALSE}
import::from(dplyr, filter)
```

  
```{r}
surveys_complete <- surveys %>%
 		filter(species_id != "",       # remove missing species_id
         !is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         sex != "")   
```
---

# Plotting time series data


```{r}
yearly_counts <- surveys_complete %>%
                 group_by(year, species_id) %>%
                 tally
```


---

## Timelapse data can be visualized as a line plot with years on the x axis and counts on the y axis:

```{r fig.height = 4}
ggplot(data = yearly_counts, aes(x = year, y = n)) +
     geom_line()
```
---


# Grouping variables in R

```{r fig.height = 4}
ggplot(data = yearly_counts, aes(x = year, y = n, group = species_id)) +
  geom_line()
```

---

# Group and add a color

```{r fig.height=4}
ggplot(data = yearly_counts, aes(
  x = year,
  y = n,
  group = species_id,
  colour = species_id
  )) +
  geom_line()
```

---

# Facet plots

```{r fig.height = 4, eval = FALSE}
ggplot(data = yearly_counts, aes(
  x = year,
  y = n,
  group = species_id,
  colour = species_id
  )) +
  geom_line() +
 facet_wrap( ~ species_id)
```

---

```{r fig.height = 6, echo = FALSE, message = FALSE}
ggplot(data = yearly_counts, aes(
  x = year,
  y = n,
  group = species_id,
  colour = species_id
  )) +
  geom_line() +
 facet_wrap( ~ species_id)
```

---

# ggplot2 has themes

The complete list of themes is available at http://docs.ggplot2.org/current/ggtheme.html. `theme_minimal()` and `theme_light()` are popular, and theme_void() can be useful as a starting point to create a new hand-crafted theme.

---

## ggthemes

```r
install.packages("ggthemes")
```

---

## Revisiting an old plot

```{r}
climate_plot <- ggplot(climate, aes(Year, Anomaly10y)) +
  geom_ribbon(
  aes(ymin = Anomaly10y - Unc10y,
  ymax = Anomaly10y + Unc10y),
  fill = "blue",
  alpha = .1
  ) +
  geom_line(color = "steelblue")
```


---
# Add an Excel 2004 theme 🙀

```{r, fig.height = 3}
library(ggthemes)
climate_plot + theme_excel()
```

---

# An economist article?

```{r, fig.height = 3}
climate_plot + theme_economist()
```

---
class: middle
# What about something nice?

```{r, fig.height = 4}
ggplot(mtcars, aes(
  x = wt,
  y = mpg,
  colour = factor(gear)
  )) +
  geom_point() +
  ggtitle("Cars") + geom_smooth(method = "lm", se = FALSE) +
  {{ scale_color_fivethirtyeight("cyl") }} +
  {{ theme_fivethirtyeight() }}
```


---

## How to save these plots

```{r, warning = FALSE}
myplt <- ggplot(mtcars, aes(
  x = wt,
  y = mpg,
  colour = factor(gear)
  )) +
  geom_point() +
  ggtitle("Cars") + geom_smooth(method = "lm", se = FALSE) +
  scale_color_fivethirtyeight("cyl") +
  theme_fivethirtyeight() 

{{ ggsave(filename = "myplot.png", width = 6, height = 4) }}
```


