---
title: "Basic statistics in R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**There is a presentation which goes with this handout, use it to add notes if you need to.**

Before we start, remember we need to do a couple of things...

#### 1. Set the working directory
See the introduction handout if you can't remember how.

Check it worked using:
```{r}
list.files()
```

#### 2. Load the packages needed
We already downloaded these and don't need to do that again. 
```{r}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(ggfortify)
```

#### 3. Import the data
We are going to work with the dataset `plant.growth.rate.csv`
```{r}
growth <- read.csv("plant.growth.rate.csv")
```

#### 4. Look at the data
```{r}
glimpse(growth)
```

### Which statistic do I need?
Before you start think very carefully about what your *question* is, why it is interesting, and how you'd plot it on a graph. If you can plot something, you can usually do the statistics!

As a (very) vague rule of thumb: 

1. If you have a continuous response (Y) variable, and a categorical predictor/explanatory (X) variable you'll need a t-test (two groups), or analysis of variance (ANOVA; two or more groups). You'd plot this as a boxplot (or barplot if you really want to). 

2. If you have a continuous response (Y) variable, and a continuous predictor/explanatory (X) variable you'll need a regression (one X), or multiple regression (two or more Xs). You'd plot this as a scatter plot.

3. If you have a continuous response (Y) variable, and a continuous predictor/explanatory (X) variable AND a categorical explanatory variable you'll need an analysis of covariance (ANCOVA). You'd plot this as a scatter plot with two or more lines (one for each category of X).

There are many other options though, this is not exhaustive and isn't always the best statistic to use for a given problem/dataset.

### How do we do statistics - step by step

1. PLOT your data
2. Fit your model
3. Look at the assumptions
4. Look at the results
5. Add interpretation to your graph

### General linear models - regression
First we'll use the plant growth dataset to demonstrate a linear regression...

1. Plot the data
Make a scatter plot of `soil.moisture.content` on the x axis and `plant.growth.rate` on the y axis.
```{r}
ggplot(dd, aes(x = soil.moisture.content, 
               y = plant.growth.rate)) + 
	geom_point(col = "cornflowerblue", size = 3) +
  labs(x = "Soil Moisture", y = "Growth Rate (mm/week)") +      theme_bw()
```

2. Make a model
```{r}
model1 <- lm(plant.growth.rate ~ soil.moisture.content, data = dd)
```

# diagnostics with ggfortify
```{r}
autoplot(mod1, smooth.colour = NA)
```

# Anova Table
```{r}
anova(mod1)
```
```{r}
summary(mod1)
```

```{r}
# ONLY with a single variable regression
ggplot(dd, aes(x = soil.moisture.content, y = plant.growth.rate)) + 
	geom_point(col = "cornflowerblue", size = 3) + geom_smooth(method = "lm", 
	se = FALSE)
```

```{r}
newxdata <- expand.grid(soil.moisture.content=seq(0.249, 1.99, length=100))
p1 <- predict(mod1, newdata = newxdata, interval="prediction")
preds <- cbind(p1, newxdata)

p2 <- predict(mod2, newdata = newxdata, interval="prediction")
preds2 <- cbind(p2, newxdata)


ggplot(data=preds, aes(x=soil.moisture.content,
                  y=fit)) +
  geom_line() +
  geom_smooth(aes(ymin=lwr, ymax=upr), stat="identity") +
  geom_point(data=dd, mapping=aes(x = soil.moisture.content,
                                  y = plant.growth.rate)) +
  geom_line(data=preds2, aes(x=soil.moisture.content,
                             y=fit), colour="red")


ggplot(dd, aes(x = soil.moisture.content, y = plant.growth.rate)) + 
  geom_point(col = "cornflowerblue", size = 3) +
  geom_smooth(method = "lm")


#  lets make some new values of soil moisture 
# so we can generate predictions from the model
# note to self: you must use the explanatory variable NAME
# that's the one on the x-axis...
newX <- expand.grid(soil.moisture.content = seq(from = 0.2, 
	to = 2, length = 10))

# look at it
newX

# Step 2
# lets get the prediction (line) and the confidence 
# bands using predict()
newY <- predict(mod1, newdata = newX, interval = "confidence")

# look at what it produces
newY

# Step 3: Housekeeping

# housekeeping (brush the two data frames together)
addThese <- data.frame(newX, newY)
addThese

# make sure names match: change fit to plant.growth.rate
addThese <- rename(addThese, plant.growth.rate = fit)
addThese

# Step 4	
# layers of fun
ggplot(dd, aes(x = soil.moisture.content, y = plant.growth.rate)) + 
	geom_point(size = 5, col = "cornflowerblue") + geom_smooth(data = addThese, 
	aes(ymin = lwr, ymax = upr), stat = "identity")

#----------------------------------------------------
# general linear models - ANOVA - daphniagrowth.csv
#----------------------------------------------------
rm(list = ls())

## read in the data
dd <- read.csv("Daphniagrowth.csv")

## look at the structure of the data
str(dd)
tbl_df(dd)


# plot
ggplot(dd, aes(x = parasite, y = growth.rate)) +
  geom_boxplot() +
	theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 15))



# summarise the data by parasite
growth.stats <- 
dd %>% 
	group_by(parasite) %>% 
		summarise(
			meanGR = mean(growth.rate), 
			seGR = sd(growth.rate)/sqrt(n()), 
			lwr = meanGR - seGR, 
			upr = meanGR + seGR)

# create the limits aesthetics for error bars
limits <- aes(ymin = lwr, ymax = upr)

# construct a fancy picture 
# note that we use the growth stats first,
# then add the raw data....

ggplot(growth.stats, aes(x = parasite, y = meanGR)) + 
	#first add the points from the growth stats"
	geom_point(colour = "red", size = 8, alpha = 0.5) + 
	
	# now add the error bars from the limits"
	geom_errorbar(limits, width = 0.1, col = "red", alpha = 0.8) + 
	
	# now add the raw data from the original dataframe
	geom_point(data = dd, aes(x = parasite, y = growth.rate), size = 3, 
		colour = "darkgreen", alpha = 0.3) + 

	# extras
	theme(axis.text.x = element_text(angle = 90, vjust = 1)) + 
	coord_flip()

# BUILD THE MODEL
mod2 <- lm(growth.rate ~ parasite, data = dd)

# diagnostics
autoplot(mod2, smooth.colour = NA)

## model interpretation
anova(mod2) ## the parasite treatment explains a large amount of variation in daphnia growth rate

summary(mm) ## Shows the effects of the parasites on growth rate, relative to the control (Intercept)

# --------------------------------------------------
# RELEVELING when reference is not alphabetical
# --------------------------------------------------

## note that the parasite variable is a factor, so lm will automatically do anova:
ggplot(dd, aes(x = parasite, y = growth.rate)) + geom_boxplot()

# to set a level, before the model
dd <- mutate(dd, parasite = relevel(parasite, ref = "Pasteuria ramosa"))
levels(dd$parasite)

# new plot with new levels
ggplot(dd, aes(x = parasite, y = growth.rate)) + geom_boxplot()
# new model with new levels
summary(lm(growth.rate ~ parasite, data = dd))

# reset to control
dd <- mutate(dd, parasite = relevel(parasite, ref = "control"))

## --------------------- EXTRAS ----------------------------------------------

# How to fit a Tukey or other post-hoc comparison using the multcomp library
mod2.aov <- aov(mod2) # repackage the model for a Tukey Test
tukey.out <- TukeyHSD(mod2.aov) # generate tukey test
tukey.out # table

# plot, setting margins and 1 x 1 grid and rotated axis labels
par(mfrow = c(1, 1), mar = c(5, 20, 4, 2))
plot(tukey.out, las = 2)

# --------------
library(multcomp)
tukey.par <- glht(mm, linfct = mcp(parasite = "Tukey"))
summary(tukey.par)
par(mfrow = c(1, 1), mar = c(5, 20, 4, 2))
plot(tukey.par)

# RESET PLOT MARGINS
par(mar = c(5, 4, 4, 2))

## Set up orthogonal (planned) contrasts: rather than ALL pairwise....
## some specific examples
c1 <- c(-1, 1/3, 1/3, 1/3) # control vs overall parasite mean
c2 <- c(0, -1, 1/2, 1/2) # parasite 1 vs 2 & 3
c3 <- c(0, 0, -1/2, 1/2) # parasite 2 vs 3

# combine the hypotheses
mat <- cbind(c1, c2, c3)

# allocate them to the treatment
contrasts(dd$parasite) <- mat

# run the model returning THESE hypotheses
mmcheck <- lm(growth.rate ~ parasite, data = dd)
summary(mmcheck)

# now set up a contrast to compare copntrol to mean of all 3 parasites
levels(dd$parasite)
contrastmatrix <- cbind(c(0, 1, 1, 1))
contrasts(dd$parasite, how.many = 1) <- contrastmatrix

# rerun
mod2 <- lm(growth.rate ~ parasite, data = dd)
summary(mod2)


#----------------------------------------------------
# general linear models - two way ANOVA - growth.csv
#----------------------------------------------------
rm(list = ls())

dd <- read.csv("growth.csv")

# alter reference level NOW to control
dd <- mutate(dd, supplement = relevel(supplement, ref = "control"))

# raw data
p1 <- ggplot(dd, aes(x = supplement, y = gain, colour = diet)) + 
	geom_point()

# the interaction plot via ggplot step 1
cows <- dd %>%
	group_by(diet, supplement) %>%
		summarise(growth = mean(gain))

# Interaction Plot via ggplot step 2
# note group argument to make sure lines connect what you want connected

p2 <- ggplot(cows, aes(x = supplement, y = growth, group = diet, colour = diet)) + 
	geom_point() + 
	geom_line()

# both on the same picture (needs gridExtra)
grid.arrange(p1, p2, ncol = 2)

# BASE functions
with(dd, interaction.plot(supplement, diet, gain))

## and the model will have the control treatment as the intercept.
## Note that we use * to fit the main effects and the interaction term

mm <- lm(gain ~ supplement * diet, data = dd)
# mm <- lm(gain ~ supplement + diet + supplemenet:diet, data = dd)

## model diagnostics
autoplot(mm, smooth.colour = NA)

## There is little evidence of an interaction, but strong main effects
anova(mm)

## Again, no clear interactions terms (there should not be,
## given the lack of a significant interaction term by F-test) 
summary(mm)
## The table looks rather scary, but is explained in the powerpoint presentation.

#----------------------------------------------------
# general linear models - ANCOVA
#----------------------------------------------------

# Get the data
rm(list = ls())
limp <- read.csv("limpet.csv")

str(limp)

# note use of scale_colour_manual
ggplot(limp, aes(x = DENSITY, y = EGGS, colour = SEASON)) + 
	geom_point(size = 5) + scale_colour_manual(values = c("yellowgreen", 
	"salmon3")) + theme_bw()

# make the model
fit2 <- lm(EGGS ~ DENSITY * SEASON, data = limp) # linear model with interacti
fit2 <- lm(EGGS ~ DENSITY + SEASON + SEASON:DENSITY, data = limp) # linear model with interacti

# check the assumptions
autoplot(fit2, smooth.colour=NA)

# examine the results
anova(fit2) # ANOVA tables with sequential sums of squares
summary(fit2) # coefficients

#----------------------------------------------------#----------------------------------------------------
# STEP 1
# construct a data frame of where you want predicted values 
# to be "located" on your graph
# play around with this to see what happens when you 
# fix DENSITY to its mean, etc....
#----------------------------------------------------
newX <- expand.grid(DENSITY = seq(from = 8, to = 50, length = 10), 
	SEASON = levels(limp$SEASON))
newX

# STEP 2
# get the predictions at the values and levels 
# specified above
newY <- predict(fit2, newdata = newX, interval = "confidence")
newY

# Step 3
# Housekeeping
PlotThis <- data.frame(newX, newY)

# Step 4 + extras
# Now use geom_smooth() to make a lovely visualisation
# of your raw data and the fitted values from the model
# and the 95% CI band around them
ggplot(limp, aes(x = DENSITY, y = EGGS, colour = SEASON)) + 
	geom_point(size = 5) +
  geom_smooth(data = filter(PlotThis, SEASON=="spring"),
              aes(y=fit, ymin = lwr, ymax = upr), stat = "identity") 









+ ....e_m_p_t_y.... = 0
# If you want to set custom colours
# do so for the raw data (colour) and the bands (fill)
scale_colour_manual(values = c("darkgreen", "hotpink")) + scale_fill_manual(values = c("darkgreen", 
	"hotpink")) + ....e_m_p_t_y.... = 0
# adding some annotation to a graph
annotate("text", x = 40, y = 2.88, label = "Hello") + ....e_m_p_t_y.... = 0
# using maths expressions on the x axis
# see ?plotmath
xlab(expression(paste(bar(x), phi))) + theme_bw()
#-----------------------------------------------
# Getting specific predictions
#-----------------------------------------------

newX <- expand.grid(DENSITY = 40, SEASON = "spring")
# get the predictions at the values and levels 
# specified above
newY <- predict(fit2, newdata = newX, interval = "confidence")
# housekeeping
data.frame(newX, newY)

#-----------------------------------------------
# t-test practical
#-----------------------------------------------

# read the data and explore
rm(list=ls())
tt <- read.csv("GardenOzone.csv")
str(tt)
tbl_df(tt)

# Make a helpful plot linked to the question about two means
ggplot(tt, aes(x = Ozone)) +
  geom_histogram(bins=5) +
  facet_wrap(~Garden.location, ncol = 1)

ggplot(tt, aes(x = Ozone)) +
  geom_dotplot(bins = 10) +
  facet_wrap(~Garden.location, ncol=1) 


# check the means and some assumptions about variance
summarise(group_by(tt, Garden.location),
          meanOzone = mean(Ozone), 
	varOzone = var(Ozone))

# Model Formulae syntax for t-test with a data frame
# note default is welch and unequal variance is OK
t.test(Ozone ~ Garden.location,
       data = tt,
       var.equal = FALSE)


ggplot(tt, aes(x = Garden.location,
               y = Ozone)) +
  geom_boxplot() +
  geom_point(position=position_jitter(width=0.05)) +
  coord_flip()
  



#--------------------------------------------------
# chi-square (Hadleyverse version)
#--------------------------------------------------

## read in the data

dd <- read.csv("ladybirds.csv")
glimpse(dd)
str(dd)

## get the number of each colour in each habitat type
lb <- group_by(dd, Habitat, colour) %>%
  summarise(obs = sum(number))

lb # LOOK AT IT.

# make a plot

ggplot(lb, aes(x = colour, y = obs, fill = colour)) +
  geom_bar(stat = "identity") + 
	facet_wrap(~Habitat) +
  scale_fill_manual(values = c(black="black", red="red"))

# the contingency table
lb_matrix <- xtabs(obs ~ Habitat + colour, data = lb)

# the test
chisq.test(lb_matrix)



# ---------------------------------------------
# USING BASE GRAPHICS
# remake the original plot
# ---------------------------------------------

par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
plot(EGGS ~ DENSITY, data = limp, cex = 2, pch = 19, col = c("green", 
	"red")[SEASON], xlab = list("DENSITY", cex = 1.3), ylab = list("EGGS", 
	cex = 1.3), ylim = c(0.2, 3), xlim = c(0, 60))

# NOW add the lines/points you made - each one of these 
# grabs a subset of the PlotThis data corresponding to SEASON 

# Summer
lines(EGGS ~ DENSITY, data = subset(PlotThis, SEASON == "summer"), 
	col = "red")

lines(lwr ~ DENSITY, data = subset(PlotThis, SEASON == "summer"), 
	col = "red", lty = 2)

lines(upr ~ DENSITY, data = subset(PlotThis, SEASON == "summer"), 
	col = "red", lty = 2)

# Spring
lines(EGGS ~ DENSITY, data = subset(PlotThis, SEASON == "spring"), 
	col = "green")

lines(lwr ~ DENSITY, data = subset(PlotThis, SEASON == "spring"), 
	col = "green", lty = 2)

lines(upr ~ DENSITY, data = subset(PlotThis, SEASON == "spring"), 
	col = "green", lty = 2)

legend(40, 3, c("Spring", "Summer"), col = c("green", "red"), 
	pch = 19)

# Making transparent confidence bands in base graphics
# note refernce to lengths in PlotThis
xx <- c(seq(0, 60, length = 10), seq(60, 0, length = 10))
yy <- c(PlotThis$upr[1:10], rev(PlotThis$lwr[1:10]))
yy2 <- c(PlotThis$upr[11:20], rev(PlotThis$lwr[11:20]))

polygon(xx, yy, col = rgb(0.19, 0.19, 0.19, 0.3), border = FALSE)
polygon(xx, yy2, col = rgb(0.19, 0.19, 0.19, 0.3), border = FALSE)


