---
title: "Dealing with data - `tidyr` and `lubridate`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**There is a presentation which goes with this handout, use it to add notes if you need to.**

Before we start, remember we need to do a couple of things...

#### 1. Set the working directory
See the introduction handout if you can't remember how.

Check it worked using:
```{r}
list.files()
```

#### 2. Load the packages needed
We already downloaded these and don't need to do that again. 
```{r}
library(tidyverse)
```

#### 3. Import the data
```{r}
museum1 <- read_csv("museum1.csv")
museum2 <- read_csv("museum2.csv")
```

### Dealing with messy data: `tidyr`
Hopefully we've taught you how to avoid making your own messy datasets! However, messy data often comes from other people (OPD - other people's data!), i.e. collaborators, supervisors, previous students of your supervisor, field technicians etc. To deal with this we can use `tidyr` and other tools like `dplyr` and `lubridate`.

Let's learn a few simple ways to deal with common issues in messy data.

First lets have a look at the first museum dataset.

```{r}
head(museum1)
```

In `museum1` we have information on how many specimens for each species are found in three different locations across the museum (the mammal tower, spirit collection and warehouse). 

Why is this untidy? What do we need to change?

### Combining multiple columns for one variable: `gather`
We need to combine the `MammalTower`, `SpiritCollection` and `Warehouse` columns into one column called `Location`. We also add a new column called `Number` that will contain the numbers that were in the `MammalTower`, `SpiritCollection` and `Warehouse` columns. 

```{r}
museum1 %>%
  gather(Location, Number, MammalTower:Warehouse)
```

Remember we need to assign the dataframe to a name if we want to use it.

```{r}
tidy.museum <-
  museum1 %>%
    gather(Location, Number, MammalTower:Warehouse)
```

### Splitting one column into many columns: `spread`

What if you need to undo this `gather` and put these variables into different columns again? We can use the `tidyr` function `spread`. This is mostly useful for multivariate statistics (e.g. principal components etc).

```{r}
tidy.museum %>%
  spread(Location, Number)
```

#### PRACTICAL

Tidy `museum1` and arrange with `Species` in reverse alphabetical order (clue - this will require a `dplyr` function).

```{r}
museum1 %>%
  gather(Location, Number, MammalTower:Warehouse) %>%
  arrange(desc(Species))
```


Next we'll play with `museum2` which is messy in some different ways. Let's have a look at the dataset.

```{r}
head(museum2)
```

In `museum2` we have information on a number of specimens. Each has an associated species name and collection date. We also have information on the sex and age of the specimen.

Why is this untidy? What do we need to change? There are several problems.

### Dealing with multiple variables in one column: `separate`
First, the `Status` column has both sex and age data. To tidy this we want to split these into separate columns.

```{r}
tidy.museum2 <-
  museum2 %>%
    separate(Status, c("Sex", "Age"), sep = " ")
```

`sep = " "` tells R that sex and age were separated in the `Status` column by a space. You may have things separated by `.` or `_` or something else.

We can combine columns using the function `unite`.

```{r}
tidy.museum2 %>%
  unite(Sex_Age, c(Sex, Age), sep = "_")
```

### Fixing dates and times with `lubridate`
Another issue in `museum2` are the collection dates. Dates and times are a real pain in data analysis. Luckily `lubridate` can help you fix these problems.

First to demonstrate the issue let's arrange the `museum2` by `CollectionDate`. What do you notice? 

```{r}
museum2 %>%
  arrange(CollectionDate)
```

This happens because R doens't know that we treat dates (and times) differently to normal numbers. Note that R thinks `CollectionDate` is a *character*.

```{r}
tbl_df(museum2)
```

To fix this we need to *parse* the dates with the `lubridate` function `dmy`. Note we use `dmy` because our data are in day, month, year format. If the dates were in month, date, year format, we'd use `mdy` and so on. 

First we need to load the library `lubridate`, as it isn't part of the `tidyverse`. Normally we'd do this at the start of the script.
```{r}
library(lubridate)
```

`lubridate` can cope with dates with all kinds of separators. Try these out to see how it arranges dates in a standard format.
```{r}
dmy("01/12/1998")
dmy("01-12-1998")
dmy("01.12.1998")
```

Now let's try and fix our dates. We will create a new variable called `DateCollected` and store this in a dataframe called `new.museum2`.

```{r}
new.museum2 <- 
  museum2 %>%
    mutate(DateCollected = dmy(CollectionDate)) %>%
    select(-CollectionDate)

new.museum2
```

Note that we also removed the now extraneous `CollectionDate` variable. 

See how this changes the date variable, is it still a character?

```{r}
tbl_df(new.museum2)
```

Did `lubridate` fix the sort issue?

```{r}
new.museum2 %>%
  arrange(DateCollected)
```

#### PRACTICAL

Using museum2, make a new dataframe called `tidy.museum`, tidied as much as possible using the tools you've just learned. 

```{r}
tidy.museum <- 
  museum2 %>%
    mutate(DateCollected = dmy(CollectionDate)) %>%
    separate(Status, c("Sex","Age")," ") %>%
    select(-CollectionDate)

tidy.museum
```

**Want to learn more data wrangling? Check out the Data Wrangling with dplyr and tidyr cheatsheet, in Help > Cheatsheets**